---
title: Doxygen Documentation of Source Code in R with rdoxygen
author: Martin Lysy and Clemens Schmid
license: GPL (>= 2)
tags: doxygen
summary: rdoxygen is an R package that simplifies rendering doxygen documentation for source code in R packages.
layout: post
src: 2018-10-20-doxygen-documentation-in-R-packages.md
---

[rdoxygen](https://CRAN.R-project.org/package=rdoxygen) is an R package for the creation of [doxygen](http://www.doxygen.org) documentation for source code (C/C++/Fortran/Java/etc.) in R packages (typically found in `src` or `inst/include`). It allows to trigger the rendering process automatically and effortlessly. Optionally, the doxygen output can be accessed by users of the package via R [vignettes](http://r-pkgs.had.co.nz/vignettes.html).

It's initially based on [Corentin M. Barbu's](https://github.com/cbarbu) answer to this [stackoverflow question](http://stackoverflow.com/questions/20713521/using-roxygen2-and-doxygen-on-the-same-package), was packaged by [Clemens Schmid](https://github.com/nevrome) and improved by [Martin Lysy](https://github.com/mlysy).

This post introduces rdoxygen v2.0.

## Quickstart

To use rdoxygen you need a working installation of the system program [doxygen](http://www.stack.nl/~dimitri/doxygen/download.html). You can either [install it from source](http://www.stack.nl/~dimitri/doxygen/manual/install.html) or rely on the package management system of your OS to get it. 

The R package rdoxygen can be installed from [CRAN](https://CRAN.R-project.org/package=rdoxygen) or -- the latest development version -- from [Github](https://github.com/nevrome/rdoxygen) with [devtools](https://CRAN.R-project.org/package=devtools) via

```{r eval=FALSE}
devtools::install_github("nevrome/rdoxygen")
```

To use roxygen to setup and afterwards update your doxygen documentation, you can simply run 

```{r eval=FALSE}
roxygen::doxy()
```

in your package directory. `doxy()` calls `doxy_init()` if there's no Doxyfile (doxygen configuration file) yet. Otherwise it just updates the documentation. 

The package provides a [RStudio Addin](https://rstudio.github.io/rstudioaddins/) named rdoxygenize that binds to the function `doxy()`. `doxy()` can therefore be called with a keyboard shortcut (e.g. <kbd>CTRL</kbd>+<kbd>SHIFT</kbd>+<kbd>-</kbd>). This makes the user experience comparable to [roxygen2](https://github.com/yihui/roxygen2) documentation via `devtools::document()` (usually <kbd>CTRL</kbd>+<kbd>SHIFT</kbd>+<kbd>D</kbd>). 

## A simple example

The rdoxygen package serves by itself as an example for its application. You can browse the code on [Github](https://github.com/nevrome/rdoxygen). For the application and preparation of rdoxygen we're going to look at the package's `DESCRIPTION` file and the directories `src`, `inst` and `vignettes`. 

In our example we have a single C++ function `docu_test_function()` that's made accessible to R with Rcpp. In the `src` directory of our package we find a header file `foo.h`, the respective implementation in `bar.cpp` and the automatically created `RcppExports.cpp`. 

```bash
|-- src
|   |-- bar.cpp
|   |-- foo.h
|   |-- RcppExports.cpp
```

In `foo.h` the `docu_test_function()` gets declared:

```{r, engine = 'Rcpp', eval = FALSE}
#ifndef FOOREGISTER_FOO_H
#define FOOREGISTER_FOO_H

#include <Rcpp.h>

double docu_test_function(double a, double b);

#endif //FOOREGISTER_FOO_H
```

The implementation belongs in `bar.cpp`:

```{r, engine = 'Rcpp', eval = FALSE}
#include "foo.h"

// [[Rcpp::export]]
double docu_test_function(double a, double b) {
  double c = a + b;
  return c;
}
```

It's possible to document C++ functions that are exposed to R with [roxygen2](https://github.com/yihui/roxygen2). To do so, we could modify the `.cpp` file and add the relevant roxygen2 tags as described in Section 2.6 of the [Rcpp Attributes vignette](https://cran.r-project.org/web/packages/Rcpp/vignettes/Rcpp-attributes.pdf).

```{r, engine = 'Rcpp', eval = FALSE}
#include "foo.h"

//' Example function: roxygen2 documentation
//' 
//' This is the roxygen2 documentation of an example function.
//' 
//' @param a A double 
//' @param b A double 
//' 
//' @return A double a + b
//' 
//' @export 
//' 
// [[Rcpp::export]]
double docu_test_function(double a, double b) {
  double c = a + b;
  return c;
}
```

Now for the idea behind rdoxygen: If we want to add doxygen documentation for this function, we can modify the `.h` file and add doxygen [comment blocks](https://www.stack.nl/~dimitri/doxygen/manual/docblocks.html):

```{r, engine = 'Rcpp', eval = FALSE}
#ifndef FOOREGISTER_FOO_H
#define FOOREGISTER_FOO_H

#include <Rcpp.h>

/**
 * \file
 *
 * Example function: doxygen documentation
 *
 * This is the doxygen documentation of an example function.
 */
double docu_test_function(double a, double b);

#endif //FOOREGISTER_FOO_H
```

`docu_test_function()` behaves documentation wise as a global function. Therefore we have to add the `\file` tag in the documentation. This is more straight forward for classes and class methods. That's where doxygen truly shines. 

If you want to add some general information at the doxygen startpage you can add it with the `\mainpage` tag anywhere in your header files: 

```{r, engine = 'Rcpp', eval = FALSE}
/*! \mainpage rdoxygen example index page
 *
 * rdoxygen test page
 * 
 * A test function is documented here: \link foo.h \endlink
 */
```

Now that we have some doxygen documentation we can set up doxygen for our R package. To do this we can rely on the the main function of rdoxygen: `doxy()`. It requires doxygen to be installed (see *Quickstart*) and to be included in the system path variable. `doxy()` will first create a Doxyfile at `inst/doc/doxygen/Doxyfile` with `doxy_init()` if it doesn't yet exist. Next, it runs doxygen on the Doxyfile and creates a directory `inst/doc/html` where you can watch the result by opening the index.hml file with a browser. If vignette = TRUE, it creates a vignette allowing the Doxygen documentation to be viewed from within R with a call to `vignette()`. More about this in the next section.

Whenever you edit the documentation or the Doxyfile you can and should rerun `doxy()` to update the documentation. To simplify this, rdoxygen provides the [RStudio Addin](https://rstudio.github.io/rstudioaddins/) *rdoxygenize*. `doxy()` can therefore be called with a keyboard shortcut.

## Access doxygen documentation in a vignette

TODO

## Background information

### Design Considerations

rdoxygen was

- *Simplicity.* A typical **rdoxygen** workflow -- heavily influenced by the design of [**devtools**](https://github.com/r-lib/devtools) -- might proceed as follows. Assuming that the package's C/C++/etc. code has been marked up for Doxygen processing, and the R working directory is *any subfolder* of the package root:

```{r, eval = FALSE}
# create default Doxyfile, process it with Doxygen, optionally wrap in R vignette
doxy(vignette = TRUE)

# separate steps above
doxy_init() # create default Doxyfile
doxy(vignette = FALSE) # process
doxy_vignette() # wrap in R vignette

# can also edit an existing Doxyfile
doxy_edit(options = c(AUTOLINK_SUPPORT = "NO"))
```

- *Stability.* The output of **rdoxygen** should function as expected with minimal additional intervention from the user. If using option `vignette = TRUE`, the package should get automatically configured to process [**rmarkdown**](https://rmarkdown.rstudio.com/) vignettes.

- *Dependencies.* Since **rdoxygen** is used for package development, it tries as much as possible to avoid adding unnecessary dependencies to the user's package (i.e., if I want to use **rdoxygen** to add Doxygen documentation to my package, ideally that shouldn't force my package users to install **rdoxygen** itself). Nevertheless helpful packages for achieving the above are:

    * [**rprojroot**](https://CRAN.R-project.org/package=rprojroot), which allows you to find the root of a package directory from any of its subfolders.
    
    * [**desc**](https://CRAN.R-project.org/package=desc), which allows you to easily edit the package `DESCRIPTION` file (e.g., for adding libraries required to process package vignettes).
    
### Arguments to Exported Functions

```{r, eval = FALSE}
# do all steps below and trigger doxygen rendering
doxy <- function(
  pkg = ".", # same as devtools::{document/load_all/install} argument, i.e., any subfolder of package root
  doxyfile = "inst/doc/doxygen/Doxyfile", # path to doxyfile relative to package root
  options, # passed to doxy_edit
  vignette = FALSE # add vignette: if TRUE then doxy_vignette is triggered
  name = "DoxygenVignette.Rmd", # passed to doxy_vignette
  index # passed to doxy_vignette
)

# add Doxyfile in package if it does not exist
doxy_init <- function(
  pkg = ".",
  doxyfile = "inst/doc/doxygen/Doxyfile"
)

# edit Doxyfile tags if doxyfile exists
doxy_edit <- function(
  pkg = ".", 
  doxyfile = "inst/doc/doxygen/Doxyfile", 
  options # named vector of key-value pairs to edit Doxyfile tags
)

# wrap Doxygen documentation in R vignette
doxy_vignette <- function(
  pkg = ".",
  name = "DoxygenVignette.Rmd", # name of Doxygen vignette
  index # name of vignette Index Entry. defaults to "C++ library documentation for package PackageName"
  overwrite = FALSE # should an existing vignette file be overwritten
)
```

### Default File Locations

* *Doxygen documentation.* Installed R vignettes can only display HTML files stored in a subfolder of `inst/doc` (as documented [here](https://github.com/nevrome/rdoxygen/issues/2#issuecomment-412536748)). Therefore, the suggested location for Doxygen documentation is `inst/doc/doxygen`.

* *Doxyfile.* As a package developer, I feel like whatever is required to create the package exactly as it should be installed on disk should be part of the package itself. In this sense, the `Doxyfile` needed to format the Doxygen documentation exactly as I want it should be part of the package as well. An obvious location for this file is `inst/doc/doxygen`. However, `inst/doc` is typically `.gitignore`d (e.g., by `devtools::use_vignette()` and `usethis::use_vignettes()`). So we'd have to manually exclude `inst/doc/doxygen/Doxyfile`, via appending an existing `.gitignore` with something like [this](https://stackoverflow.com/questions/5533050/gitignore-exclude-folder-but-include-specific-subfolder):
    
```
# paste the following to the bottom of existing .gitignore

!inst/doc # unignores inst/doc
inst/doc/* # ignore everything inside inst/doc but not inst/doc itself
!inst/doc/doxygen/Doxyfile # unignore Doxyfile
```
   
### Default Doxyfile Tags

First, it should be noted that the `INPUT` tag cannot handle relative directories outside of where it's run, i.e.:

```
# in Doxyfile
INPUT = src/ # works fine
INPUT = ../src/ # does not work
```

Thus, **rdoxygen** runs Doxygen *from the package root folder*. That being said, here are the default tags rdoxygen sets in the Doxyfile:

```
INPUT = src/ inst/include/ # the two locations in which you expect to find C++ code
OUTPUT_DIRECTORY = inst/doc/doxygen
PROJECT_NAME = "C++ Library Documentation for Package PackageName"
```

Also note that another useful option might be `USE_MATHJAX = YES`, which makes formulas look much nicer than when this option is set to `NO`.
